Language Attitudes of Catalan-Spanish Bilinguals Towards Intervocalic Alveolar Fricatives in Catalan

### 2024-05-03 manuscript/main.tex

Catalan and Spanish are two languages that have coexisted for centuries,
mainly in eastern Spain, along the Mediterranean coast \textbf{(Hualde,
1992)}. It is safe to assume that there is a bidirectional influence
between these languages: in that region, Catalan has influenced Spanish
and Spanish has influenced Catalan in all linguistic levels
\textbf{(Davidson, 2014, 2015)}. A crucial area of cross-linguistic
influence might be the production of sounds, especially in those sounds
that have different distribution in Catalan and Spanish. For example,
previous studies have found that, in Barcelona, the production of
Catalan vowels is different based on the L1 of the speaker, being it
less target-like in L1 Spanish speakers compared to L1 Catalan speakers
\textbf{(CITAR SEBASTIAN GALLES, etc)}.

However, little is known about the perception and the attitudes toward a
less Catalan-like production of a given sound. Therefore, the aim of
this study is to examine the covert attitudes towards the production of
intervocalic /s/, as either {[}s{]} (disfavored by prescriptivist norms)
or {[}z{]} (preferred by prescriptivist norms) by L1 Catalan, L1
Spanish, and Catalan-Spanish bilinguals from Catalonia using a matched
guise test \textbf{(Lambert, 1967)}. The alveolar fricative was chosen
because it exists in both Catalan and Spanish inventories, but their
contextual distribution is different \textbf{(Llisterri, 2023)}.

@@

Both Catalan and Peninsular Spanish sound inventories have a phoneme
/s/, which can be classified as a voiceless alveolar fricative
\cite{hualde_colina_2015}. In addition, most of the Catalan varieties
have a voiced alveolar phoneme in intervocalic contexts \textbf{(Lloret,
2011; Recasens Vives, 1996, p.~279)}. In onset position, /s/ and /z/
contrast phonemically, which leads to minimal pairs like \emph{cel}
{[}s\textepsilon l{]} - \emph{zel} {[}z\textepsilon l{]} and
\emph{fossa} {[}fósa{]} - \emph{fosa} {[}fóza{]} \textbf{(Carrera-Sabaté
\& Prieto, 2012-2019)}. Conversely, at the end of the word, this
contrast is neutralized, being the voicing or devoicing of the alveolar
fricative dependent on the following sound \textbf{(Institut d'Estudis
Catalans, 2022)}. For example, if it is followed by a vowel, the
alveolar fricative is voiced: \emph{torres} {[}tór\textschwa{}s{]} and
\emph{torres antigues}
{[}tor\textschwa{}z\textschwa{}nti\textgamma{}\textschwa{}s{]}.

\textbackslash cite\{hualde\_colina\_2015

While Spanish inventory does not have the phoneme /z/, the sound {[}z{]}
may occur as an allophone of /s/, due to a voice assimilation process.
It is important to mention, though, that it has been claimed that this
voicing is generally not complete, but partial \textbf{(Benet et al.,
2012; Hualde \& Colina, 2015; Hualde et al., 2020)}. There is some
evidence of the influence of Spanish on the production of the Catalan
alveolar fricative, in contexts where, according to prescriptivist rules
\textbf{(Institut d'Estudis Catalans, 2022)}, the {[}z{]} should be
produced, but {[}s{]} is produced instead. This phenomenon has mainly
been attested in those areas where there has been an extended contact
between Catalan and Spanish, such as the so-called ``franja de Aragón''
(borderland between Catalonia and Aragon), some areas of the Valencian
Community, and Barcelona and other urban areas \textbf{(Benet et al.,
2012; Davidson, 2020; Recasens Vives, 1996, p.~268)}. The opposite
pattern has also been observed - that is, the voicing of the alveolar
fricative in the Spanish spoken in Barcelona and Valencia.
Interestingly, although in both Barcelona and Valencia Catalan and
Spanish coexist and influence each other, it seems that Catalan has a
greater influence on Spanish in Barcelona, but Spanish has a greater
influence on Catalan in Valencia \textbf{(Davidson, 2020)}.

@@

The social meaning and the perception of intervocalic {[}z{]} in Spanish
has been studied in a variety of contexts, including both monolingual
\textbf{(Chappell, 2016, 2017, 2021; Chappell et al., 2023; Chappell \&
García, 2017; García, 2015, 2019; Silverstein, 2003)} and bilingual
contexts \textbf{(Davidson, 2014, 2015, 2019; Sinner, 2002)}, despite
having been categorized as a low-salient variant \textbf{(Chappell,
2017, 2021; Chappell et al., 2023; García, 2015)}. With respect to the
former, it has been found that for listeners from Costa Rica and Ecuador
(García, 2019), the covert attitudes associated with intervocalic
{[}z{]} vary significantly by speaker gender. Consequently, male
speakers who use {[}z{]} tend to be positively evaluated, and generally
perceived as nice, local, confident, and masculine. Conversely, female
speakers who use {[}z{]} are only afforded negative associations of
lower education and social class \textbf{(Chappell, 2016; Chappell et
al., 2023; Silverstein, 2003)}. Additionally, this variable has been
found to serve as a marker of regional identity, with voicing being
associated with neighboring Highland Ecuadorian dialects that exhibit
systematic voicing.

With respect to the social meaning and the perception of intervocalic
{[}z{]} in bilingual contexts, this phenomenon has been mainly studied
in Catalan-Spanish contact situations \textbf{(Davidson, 2014, 2015)}.
According to \textbf{Sinner (2002)}, intervocalic {[}z{]} can be
considered a possible linguistic marker of Catalonian Spanish, despite
its relative lack of social awareness \textbf{(Davidson, 2019; Sinner,
2002)}. Additionally, Davidson (2019) found that, in Barcelona, the
indexical field of intervocalic {[}z{]} features exclusively positive
associations with bilingualism and local solidarity, which warrants its
gradual adoption in the speech community, unlike other phonetic markers
of Catalonian Spanish that are the topic of overt social ridicule, such
as lateral velarization. These results are in line with previous
findings on intervocalic {[}z{]} in Catalonian Spanish. For instance,
\textbf{Silverstein (2003)} found that intervocalic {[}z{]} indexes a
local Barcelona identity, and the variant then acquires positive
associations with bilingualism and cultural pride. Overall, these
findings suggest that intervocalic {[}z{]} likely functions as a
sociolinguistic indicator of Catalonian Spanish within the Catalonian
speech community, indexing Catalan-Spanish bilingual membership while
lacking any considerable degree of generalized conscious awareness
\textbf{(Davidson, 2019)}.

@@

Data were collected using an online survey hosted on Qualtrics that
consisted of two parts. The first part, a demographic questionnaire,
collected information regarding participant age, gender, place of
origin, place of residence, mother tongue, family income, and level of
studies. The second portion of the survey consisted of a perception task
where participants were presented with a series of stimuli containing
the target phoneme contrast (intervocalic voiced and voiceless alveolar
fricative) such as \emph{han tancat les vies estretes} `they closed the
narrow streets', followed by 11 evaluative statements related to social
characteristics such as level of studies, friendliness, social class,
and ethnic origin. These evaluative statements were then followed by a
100-point draggable sliding scale with which participants indicated
their level of agreement or disagreement with the content of the
statement, where 0 was ``totally disagree'' and 100 was ``totally
agree''.

@@

The aim of this exploration study was to explore the covert attitudes
that are linked to the perception Catalan intervocalic /s/ as either
{[}s{]} or {[}z{]} and to examine the relationship between individual
social factors and covert attitudes towards Catalan intervocalic /s/.
The results suggest that these predictors have a different effect on the
ratings of the speaker depending on the social characteristic under
evaluation.

Contrary to what we expected, phoneme was not a significant predictor.
One explanation for this divergent pattern could be that either
participants could not distinguish between {[}s{]} and {[}z{]}
\textbf{(Davidson, 2019)} or, if they could, this difference in
production was not changing their attitudes towards the speaker
\textbf{(Sinner, 2002)}. This finding is interesting considering that
the production of the voiceless alveolar fricative in intervocalic
position has been considered a result of Catalan in contact with Spanish
\textbf{(Benet et al., 2012; Davidson, 2020; Recasens Vives, 1996,
p.~268)}, and, from a prescriptivist point of view, it should be avoided
\textbf{(Institut d'Estudis Catalans, 2022)}. Despite the lack of
significance of this predictor, it should be mentioned that the speaker
received more positive ratings when the {[}z{]} was being produced. The
fact that phoneme was not an important factor contrasts with
\textbf{Davidson (2020)}, who analyzed covert and overt attitudes
towards Catalonian Spanish and found that {[}z{]} was a significant
predictor: it was considered a sociolinguistic indicator of
Catalan-Spanish bilingual membership. Although at this point we can only
hypothesize, it could be that {[}z{]} is perceptually more salient in
Spanish than in Catalan, as the Spanish inventory does not have the
phoneme /z/, but voicing of /s/ might partially occur \textbf{(Benet et
al., 2012; Hualde \& Colina, 2015; Hualde et al., 2020)}. Another
explanation could be that the /s/ alone does not provide enough social
information; rather, the combination of the phoneme /s/ with the
precedent vowel might give the listener more cues and have a stronger
social meaning. Future studies should address this issue in order to
explore if the combination of vowel + /s/ carries other social
connotations than just the /s/ alone.

We expected mother tongue to be a strong predictor of the attitudes
towards the production of intervocalic /s/, specifically, that L1
Catalan and bilingual speakers would give higher rates to those
statements with {[}z{]}. This hypothesis was only partially confirmed.
As reported above, phoneme was not a factor influencing the attitudes,
and mother tongue was only a significant predictor in the two statements
related to Catalan identity, namely \emph{Parla bé català} `S/he speaks
proper Catalan' and \emph{És d'origen català} `S/he is of Catalan
origin'. Interestingly, bilinguals gave higher ratings to the speaker,
followed by L1 Catalan participants and L1 Spanish participants. A
potential explanation for this finding, that should be verified in
future research, could be that L1 Catalan speakers have specific ideas
(higher standards) towards the `proper' Catalan. In addition, some
studies have suggested that those Catalan people who grow up in
Spanish-speaking households typically exhibit more negative attitudes
towards Catalan \textbf{(CITATIONS)}, which is consistent with our
results.

Lastly, province was also found as an important predictor, but only for
some statements. In two of the solidarity statements (\emph{És simpàtic}
`S/he is nice'; \emph{És de fiar} `S/he is trustworthy'), the speaker
was rated more positively by those participants from the province of
Barcelona. This finding might suggest that they identify with the
speaker; thus, giving her higher ratings \textbf{BUSCAR CITA}.
Interestingly, the speaker that created the stimuli is from the province
of Girona, but has spent the last 11 years of her life living outside
that area, either in Barcelona or in another country. We asked her to
read the sentences using a Central Catalan `neutral accent', which
people might have potentially related with Barcelona. This `neutral
accent' does not show any of the prototypical characteristics of the
\emph{accent gironí}, a subvariety within Central Catalan spoken in some
areas of Girona \textbf{(Gros, 2024)}.

>>>>

Catalan and Spanish are two languages that have coexisted for centuries,
mainly in eastern Spain, along the Mediterranean coast
\cite{hualde_1992} It is safe to assume that there is a bidirectional
influence between these languages: in that region, Catalan has
influenced Spanish and Spanish has influenced Catalan in all linguistic
levels \cite{davidson_2014} \cite{davidson_2015}. A crucial area of
cross-linguistic influence might be the production of sounds, especially
in those sounds that have different distribution in Catalan and Spanish.
For example, previous studies have found that, in Barcelona, the
production of Catalan vowels is different based on the L1 of the
speaker, being it less target-like in L1 Spanish speakers compared to L1
Catalan speakers \cite{pallier_bosch_sebastian_galles_1997}.

However, little is known about the perception and the attitudes toward a
less Catalan-like production of a given sound. Therefore, the aim of
this study is to examine the covert attitudes towards the production of
intervocalic /s/, as either {[}s{]} (disfavored by prescriptivist norms)
or {[}z{]} (preferred by prescriptivist norms) by L1 Catalan, L1
Spanish, and Catalan-Spanish bilinguals from Catalonia using a matched
guise test \cite{lambert_1967}. The alveolar fricative was chosen
because it exists in both Catalan and Spanish inventories, but their
contextual distribution is different \cite{llisterri_2023}.

@@

Both Catalan and Peninsular Spanish sound inventories have a phoneme
/s/, which can be classified as a voiceless alveolar fricative
\cite{hualde_2015}. In addition, most of the Catalan varieties have a
voiced alveolar phoneme in intervocalic contexts \cite{lloret_2011}
\cite{recasens_1996}. In onset position, /s/ and /z/ contrast
phonemically, which leads to minimal pairs like \emph{cel}
{[}s\textepsilon l{]} - \emph{zel} {[}z\textepsilon l{]} and
\emph{fossa} {[}fósa{]} - \emph{fosa} {[}fóza{]}
\cite{carrera_sabate_prieto_2012_2019}. Conversely, at the end of the
word, this contrast is neutralized, being the voicing or devoicing of
the alveolar fricative dependent on the following sound \cite{iec_2022}.
For example, if it is followed by a vowel, the alveolar fricative is
voiced: \emph{torres} {[}tór\textschwa{}s{]} and \emph{torres antigues}
{[}tor\textschwa{}z\textschwa{}nti\textgamma{}\textschwa{}s{]}.

While Spanish inventory does not have the phoneme /z/, the sound {[}z{]}
may occur as an allophone of /s/, due to a voice assimilation process.
It is important to mention, though, that it has been claimed that this
voicing is generally not complete, but partial
\cite{benet_cortes_lleo_2012} \cite{hualde_2015}
\cite{hualde_et_al_2020}. There is some evidence of the influence of
Spanish on the production of the Catalan alveolar fricative, in contexts
where, according to prescriptivist rules \cite{iec_2022}, the {[}z{]}
should be produced, but {[}s{]} is produced instead. This phenomenon has
mainly been attested in those areas where there has been an extended
contact between Catalan and Spanish, such as the so-called ``franja de
Aragón'' (borderland between Catalonia and Aragon), some areas of the
Valencian Community, and Barcelona and other urban areas
\cite{benet_cortes_lleo_2012} \cite{recasens_1996} \cite{davidson_2020}.
The opposite pattern has also been observed - that is, the voicing of
the alveolar fricative in the Spanish spoken in Barcelona and Valencia.
Interestingly, although in both Barcelona and Valencia Catalan and
Spanish coexist and influence each other, it seems that Catalan has a
greater influence on Spanish in Barcelona, but Spanish has a greater
influence on Catalan in Valencia \cite{davidson_2020}.

@@

The social meaning and the perception of intervocalic {[}z{]} in Spanish
has been studied in a variety of contexts, including both monolingual
\cite{chappel_et_al_2023} \cite{chappel_2021} \cite{garcia_2019}
\cite{chappel_2016} \cite{chappell_2017} \cite{chappell_garcia_2017}
\cite{silverstein_2003} and bilingual contexts \cite{davidson_2014}
\cite{davidson_2015} \cite{davidson_2019} \cite{sinner_2002}, despite
having been categorized as a low-salient variant
\cite{chappel_et_al_2023} \cite{chappel_2021} \cite{chappell_2017} .
With respect to the former, it has been found that for listeners from
Costa Rica and Ecuador \cite{garcia_2019}, the covert attitudes
associated with intervocalic {[}z{]} vary significantly by speaker
gender. Consequently, male speakers who use {[}z{]} tend to be
positively evaluated, and generally perceived as nice, local, confident,
and masculine. Conversely, female speakers who use {[}z{]} are only
afforded negative associations of lower education and social class
\cite{chappel_et_al_2023} \cite{chappel_2016} \cite{silverstein_2003}.
Additionally, this variable has been found to serve as a marker of
regional identity, with voicing being associated with neighboring
Highland Ecuadorian dialects that exhibit systematic voicing.

With respect to the social meaning and the perception of intervocalic
{[}z{]} in bilingual contexts, this phenomenon has been mainly studied
in Catalan-Spanish contact situations \cite{davidson_2014}
\cite{davidson_2015}. According to Sinner \cite{sinner_2002},
intervocalic {[}z{]} can be considered a possible linguistic marker of
Catalonian Spanish, despite its relative lack of social awareness
\cite{davidson_2019} \cite{sinner_2002}. Additionally, Davidson (2019)
found that, in Barcelona, the indexical field of intervocalic {[}z{]}
features exclusively positive associations with bilingualism and local
solidarity, which warrants its gradual adoption in the speech community,
unlike other phonetic markers of Catalonian Spanish that are the topic
of overt social ridicule, such as lateral velarization. These results
are in line with previous findings on intervocalic {[}z{]} in Catalonian
Spanish. For instance, Silverstein \cite{silverstein_2003} found that
intervocalic {[}z{]} indexes a local Barcelona identity, and the variant
then acquires positive associations with bilingualism and cultural
pride. Overall, these findings suggest that intervocalic {[}z{]} likely
functions as a sociolinguistic indicator of Catalonian Spanish within
the Catalonian speech community, indexing Catalan-Spanish bilingual
membership while lacking any considerable degree of generalized
conscious awareness \cite{davidson_2019}.

@@

Data were collected using an online survey hosted on Qualtrics that
consisted of two parts. The first part, a demographic questionnaire,
collected information regarding participant age, gender, place of
origin, place of residence, mother tongue, family income, and level of
studies. The second portion of the survey consisted of a perception task
where participants were presented with a series of stimuli containing
the target phoneme contrast (intervocalic voiced and voiceless alveolar
fricative) such as \emph{Han tancat les vies estretes} `they closed the
narrow streets', followed by 11 evaluative statements related to social
characteristics such as level of studies, friendliness, social class,
and ethnic origin. These evaluative statements were then followed by a
100-point draggable sliding scale with which participants indicated
their level of agreement or disagreement with the content of the
statement, where 0 was ``totally disagree'' and 100 was ``totally
agree''.

@@

The aim of this exploratory study was to explore the covert attitudes
that are linked to the perception Catalan intervocalic /s/ as either
{[}s{]} or {[}z{]} and to examine the relationship between individual
social factors and covert attitudes towards Catalan intervocalic /s/.
The results suggest that these predictors have a different effect on the
ratings of the speaker depending on the social characteristic under
evaluation.

Contrary to what we expected, phoneme was not a significant predictor.
One explanation for this divergent pattern could be that either
participants could not distinguish between {[}s{]} and {[}z{]}
\cite{davidson_2019} or, if they could, this difference in production
was not changing their attitudes towards the speaker \cite{sinner_2002}.
This finding is interesting considering that the production of the
voiceless alveolar fricative in intervocalic position has been
considered a result of Catalan in contact with Spanish
\cite{benet_cortes_lleo_2012} \cite{recasens_1996} \cite{davidson_2020},
and, from a prescriptivist point of view, it should be avoided
\cite{iec_2022}. Despite the lack of significance of this predictor, it
should be mentioned that the speaker received more positive ratings when
the {[}z{]} was being produced. The fact that phoneme was not an
important factor contrasts with Davidson \cite{davidson_2020}, who
analyzed covert and overt attitudes towards Catalonian Spanish and found
that {[}z{]} was a significant predictor: it was considered a
sociolinguistic indicator of Catalan-Spanish bilingual membership.
Although at this point we can only hypothesize, it could be that {[}z{]}
is perceptually more salient in Spanish than in Catalan, as the Spanish
inventory does not have the phoneme /z/, but voicing of /s/ might
partially occur \cite{benet_cortes_lleo_2012} \cite{hualde_2015}
\cite{hualde_et_al_2020}. Another explanation could be that the /s/
alone does not provide enough social information; rather, the
combination of the phoneme /s/ with the precedent vowel might give the
listener more cues and have a stronger social meaning. Future studies
should address this issue in order to explore if the combination of
vowel + /s/ carries other social connotations than just the /s/ alone.

We expected mother tongue to be a strong predictor of the attitudes
towards the production of intervocalic /s/, specifically, that L1
Catalan and bilingual speakers would give higher rates to those
statements with {[}z{]}. This hypothesis was only partially confirmed.
As reported above, phoneme was not a factor influencing the attitudes,
and mother tongue was only a significant predictor in the two statements
related to Catalan identity, namely \emph{Parla bé català} `S/he speaks
proper Catalan' and \emph{És d'origen català} `S/he is of Catalan
origin'. Interestingly, bilinguals gave higher ratings to the speaker,
followed by L1 Catalan participants and L1 Spanish participants. A
potential explanation for this finding, that should be verified in
future research, could be that L1 Catalan speakers have specific ideas
(higher standards) towards the `proper' Catalan. In addition, some
studies have suggested that those Catalan people who grow up in
Spanish-speaking households typically exhibit more negative attitudes
towards Catalan \cite{aguilo_more_lynch_2017}, which is consistent with
our results.

Lastly, province was also found as an important predictor, but only for
some statements. In two of the solidarity statements (\emph{És simpàtic}
`S/he is nice'; \emph{És de fiar} `S/he is trustworthy'), the speaker
was rated more positively by those participants from the province of
Barcelona. This finding might suggest that they identify with the
speaker; thus, giving her higher ratings \textbf{BUSCAR CITA}.
Interestingly, the speaker that created the stimuli is from the province
of Girona, but has spent the last 11 years of her life living outside
that area, either in Barcelona or in another country. We asked her to
read the sentences using a Central Catalan `neutral accent', which
people might have potentially related with Barcelona. This `neutral
accent' does not show any of the prototypical characteristics of the
\emph{accent gironí}, a subvariety within Central Catalan spoken in some
areas of Girona \cite{gros_2024}.
